
## Exercises

1\. Suppose you poll a population in which a proportion $p$ of voters are Democrats and $1-p$ are Republicans. Your sample size is $N=25$. Consider the random variable $S$ which is the **total** number of Democrats in your sample. What is the expected value of this random variable? Hint: it's a function of $p$.


S=N*p

```{r}

```

2\. What is the standard error of $S$ ? Hint: it's a function of $p$.
$\sqrt{(Npq)}$

3\. Consider the random variable $S/N$. This is equivalent to the sample average, which we have been denoting as $\bar{X}$. What is the expected value of the $\bar{X}$? Hint: it's a function of $p$.

p

4\. What is the standard error of $\bar{X}$? Hint: it's a function of $p$.

SE=\sqrt{p(1-p)/N}

5\. Write a line of code that gives you the standard error `se` for the problem above for several values of $p$, specifically for `p <- seq(0, 1, length = 100)`. Make a plot of `se` versus `p`.

```{r}
p <- seq(0, 1, length = 100)
dat<-data.frame(p=p,
          se=sqrt(p*(1-p)/25 ))

dat%>%
  ggplot2::ggplot(aes(x=p,y=se))+
  geom_point()
```


6\. Copy the code above and put it inside a for-loop to make the plot for $N=25$, $N=100$, and $N=1000$.


```{r}
p <- seq(0, 1, length = 100)
dat<-NULL
Nlist<-c(25,100,1000)
for (n in Nlist){
  
temp<-data.frame(p=p,
          se=sqrt(p*(1-p)/n ),
          N=n)
dat<-rbind(dat,temp)
}

dat%>%
  ggplot2::ggplot(aes(x=p,y=se))+
  geom_point()+
  facet_wrap(~N)

```
7\. If we are interested in the difference in proportions, $p - (1-p)$, our estimate is $d = \bar{X} - (1-\bar{X})$. Use the rules we learned about sums of random variables and scaled random variables to derive the expected value of $d$.

d=p-(1-p)=2p-1

8\. What is the standard error of $d$?

$$
2*\sqrt{p(1-p)/N}
$$

9\. If the actual $p=.45$, it means the Republicans are winning by a relatively large margin since $d= -.1$, which is a 10% margin of victory. In this case, what is the standard error of $2\hat{X}-1$ if we take a sample of $N=25$?
```{r}

2*sqrt(0.45*0.55/N)
B=10000
poll<-function(N){
  x<-sample(c(0,1), N, replace = TRUE, prob=c(0.55, 0.45))
 2*sum(x)/N-1
}


result<-replicate(B,poll(25))
sd(result)

```

10\. Given the answer to 9, which of the following best describes your strategy of using a sample size of $N=25$?

a.  The expected value of our estimate $2\bar{X}-1$ is $d$, so our prediction will be right on.
b.  Our standard error is larger than the difference, so the chances of $2\bar{X}-1$ being positive and throwing us off were not that small. We should pick a larger sample size.
c.  The difference is 10% and the standard error is about 0.2, therefore much smaller than the difference.
d.  Because we don't know $p$, we have no way of knowing that making $N$ larger would actually improve our standard error.


B

